MS Prelim Data K23

### 2024-05-24 scripts/first_pass_adapting_K23_analysis_20231031.Rmd
Add files via upload

```{r read_in_df_and_subset}
homedir <- "/Users/eballer/BBL/msanxiety/"
columns_to_make_integer <- c("ACCESSION_NUM", "PAT_AGE_AT_EXAM","MRI_ENC_AGE","CURRENT_AGE","hemoglobin", "WBC","RBC","B12","FOLATE", "TSH", "RPR","CSFAPPEAR1","CSFAPPEAR4","CSFCOLOR1",       "CSFCOLOR4","CSFTUBE","CSFTUBE4","PHQ.2","PHQ.9") #removed "VitD", not in new pull for some reason

#open Rds
data_2023_pull <- readRDS(paste0(homedir, "/data/data_2023_pull_with_curated_icd10_codes.rds"))

#Get med info in
#read in BCP info, names of meds, est and prog and doses
bcp <- read.csv("/Users/eballer/BBL/medication_data/BCP_est_and_prog_levels_and_doses_all_brands.csv", sep = ",", header = TRUE)
bcp$bcp_name <- toupper(bcp$bcp_name)

#read in ms med info, names of meds, mechanism, and whether or not dmt
msmeds <- read.csv("/Users/eballer/BBL/medication_data/ms_medications_brand_and_generic.csv", sep = ",", header = TRUE)
msmeds$Medication <- toupper(msmeds$Medication)

#depression/psych meds from NAMI website 
psych_meds <- read.csv("/Users/eballer/BBL/medication_data/nami_psych_meds_antidepressants.csv", sep = ",", header = TRUE)
psych_meds$Medication.Name <- toupper(psych_meds$Medication.Name)

#anxiety meds from chatgpt (cross referenced with psych meds)
anxiety_meds <- read.csv("/Users/eballer/BBL/medication_data/antianxiety_medications.csv", sep = ",", header = TRUE)
anxiety_meds$Medication <- toupper(anxiety_meds$Medication)
anxiety_meds_no_beta_blocker <- anxiety_meds %>% filter(Class != "Beta_blocker")

##########################
### some preprocessing ###
##########################
#we need to keep scans from people seen by an MS provider (n = 17067) who have an MS diagnostic code (n=16,830)
#we make a bunch of columns from character to integer, binarize sex and race, and put date into a nice format YYYYMMDD, the %m%d%y indicates what it started out as

#goal is to keep people who were seen by 
 data_empi_acc_f_phq <-data_2023_pull %>% 
   mutate(across(.cols = columns_to_make_integer, .fns = as.integer)) %>%
   mutate(sex_binarized = ifelse(SEX == "MALE", 1, 2)) %>%
   mutate(osex = ordered(sex_binarized,levels = c(1,2), labels = c("Male","Female"))) %>%
   mutate(race_binarized = ifelse(RACE == "WHITE", 1, 2)) %>%
   mutate(orace = ordered(race_binarized,levels = c(1,2), labels = c("White","Non-white"))) %>%
   mutate(EXAM_DATE = as.Date(BEGIN_EXAM_DTTM, format = "%m/%d/%y")) %>% 
   mutate(EXAM_DATE = gsub(EXAM_DATE, pattern = "-", replacement = "")) %>%
   mutate(BEGIN_EXAM_DTTM = as.Date(BEGIN_EXAM_DTTM, format = "%m/%d/%y")) %>% #make sure to actually save it as date object for later processing
   mutate(EMPI = as.factor(EMPI)) %>%
   mutate(number_of_dx = ifelse(!is.na(ICD10), (str_count(ICD10, ";") + 1), 0)) %>%
   mutate(number_of_dx_extended = ifelse(!is.na(all_icd10_dx), str_count(all_icd10_dx, ";") + 1, 0)) %>%
   mutate(number_of_psychiatric_comorbidities_extended = ifelse(!is.na(all_icd10_dx), str_count(all_icd10_dx, "F"), 0)) %>%
   mutate(number_of_anxiety_comorbidities_extended = ifelse(!is.na(all_icd10_dx), str_count(all_icd10_dx, "F4"), 0)) %>%
   mutate(number_of_depression_comorbidities_extended = ifelse(!is.na(all_icd10_dx), str_count(all_icd10_dx, "F3"), 0)) %>%
   mutate(number_of_medications = ifelse(!is.na(Medication), str_count(Medication, ";") + 1, 0)) %>%
   mutate(On.Bcp = ifelse(str_detect(Medication, paste(bcp$bcp_name, collapse = "|")), 1, 0)) %>%
   mutate(On.Psych.Meds = ifelse(str_detect(Medication, paste(psych_meds$Medication, collapse = "|")), 1, 0)) %>%
   mutate(On.Antidepressants = ifelse(str_detect(Medication, paste(psych_meds$Medication[psych_meds$Antidepressant == "t"], collapse = "|")), 1, 0)) %>%
   mutate(On.Anxiolytics = ifelse(str_detect(Medication, paste(anxiety_meds$Medication, collapse = "|")), 1, 0)) %>%
   mutate(On.Anxiolytics_no_beta_blocker = ifelse(str_detect(Medication, paste(anxiety_meds_no_beta_blocker$Medication, collapse = "|")), 1, 0)) %>%
   mutate(On.Benzos = ifelse(str_detect(Medication, paste(anxiety_meds$Medication[anxiety_meds$Class == "Benzodiazepine"], collapse = "|")), 1, 0)) %>%
   mutate(On.Anti_cd20 = ifelse(str_detect(Medication, paste(msmeds$Medication[msmeds$Mechanism == "anti-cd20"], collapse = "|")), 1, 0)) %>%
   mutate(On.Interferon = ifelse(str_detect(Medication, paste(msmeds$Medication[msmeds$Mechanism == "Interferon"], collapse = "|")), 1, 0)) %>%
   mutate(On.Steroids = ifelse(str_detect(Medication, paste(msmeds$Medication[msmeds$Mechanism == "Steroid"], collapse = "|")), 1, 0)) %>%
   mutate(On.Dmt = ifelse(str_detect(Medication, paste(msmeds$Medication[msmeds$DMT == "y"], collapse = "|")), 1, 0)) %>%
   mutate(Has.PHQ2 = ifelse(PHQ.2 != "NULL", 1, 0)) %>%
   mutate(Has.PHQ9 = ifelse(PHQ.9 != "NULL", 1, 0)) %>%
   mutate(Has.depdx = ifelse(grepl("F3", all_icd10_dx), 1, 0)) %>%
   mutate(Has.anxietydx = ifelse(grepl("F4", all_icd10_dx), 1, 0)) %>%
   mutate(Has.depANDanx.dx = ifelse(Has.depdx & Has.anxietydx, 1, 0)) %>%
   mutate(PHQ.2_modsev_dep_sxs = ifelse(Has.PHQ2 & PHQ.2 >= 3, 1, 0)) %>%
   mutate(PHQ.9_modsev_dep_sxs = ifelse(Has.PHQ9 & PHQ.9 >= 10, 1, 0)) %>%
   mutate(PHQ.2_mild_dep_sxs = ifelse(Has.PHQ2 & PHQ.2 > 0 & PHQ.2 < 3, 1, 0)) %>%
   mutate(PHQ.9_mild_dep_sxs = ifelse(Has.PHQ9 & PHQ.9 > 0 & PHQ.9 < 10, 1, 0)) %>%
   mutate(PHQ.2_zero = ifelse(Has.PHQ2 & PHQ.2 == 0, 1, 0)) %>%
   mutate(PHQ.9_zero = ifelse(Has.PHQ9 & PHQ.9 == 0, 1, 0)) %>%
   mutate(dep_by_dx_phq = ifelse((Has.depdx | PHQ.2_modsev_dep_sxs | PHQ.9_modsev_dep_sxs),1,0)) %>%
   mutate(dep_by_dx_phq_antidep = ifelse((Has.depdx | On.Antidepressants | PHQ.2_modsev_dep_sxs | PHQ.9_modsev_dep_sxs),1,0)) %>%
   mutate(true_healthy = ifelse((!Has.depdx & !On.Psych.Meds & (PHQ.2_zero | PHQ.9_zero)), 1, 0)) %>%
   mutate(dep_by_dx_phq_meds_healthy_phq0_no_psych_meds = ifelse(dep_by_dx_phq_antidep | true_healthy, 1, 0)) %>%
   rowwise(ACCESSION_NUM) %>%
   mutate(depGroupVar =
            sum(c(dep_by_dx_phq_meds_healthy_phq0_no_psych_meds,dep_by_dx_phq_antidep))) %>%#you get an extra point if you are in the depression group, so the end result is dep = 2, healthy = 1, 0 for exclude
   ungroup() #n=3,737 unique people, n= 16,830 total

##########################################################
#             Read in Fascicle info             #
##########################################################

fascicle_proportions <- read.csv(paste0(homedir, "/results/fascicle_volumes_all_subjects_roi_n2336.csv"), header = T, sep = ",") #n = 2336

@@

write.table(fascicle_names,"/Users/eballer/BBL/msdepression/templates/dti/HCP_YA1065_tractography/fascicle_names.csv", row.names = F, quote = F, col.names = F)

########
#fascicle volumes in depression network
########
fascicle_volumes_dep <- read.csv(paste0(homedir, "results/streamline_volume_within_dep_network.csv"), header = T, sep = ",")

@@

write.table(fascicle_volumes_dep_network, paste0(homedir,  "/templates/dti/HCP_YA1065_tractography/fascicle_names_dep_network_grouping.csv"), sep = ",", col.names=T)#, row.names = F, quote = F, col.names = T)

@@

#fascicle_names_dep_network_all <- fascicle_volumes_dep_network$fascicle[which(fascicle_volumes_dep_network$inDepMask == 1)]
#write.table(fascicle_names_dep_network_all,"/Users/eballer/BBL/msdepression/templates/dti/HCP_YA1065_tractography/fascicle_names_dep_network_all.csv", row.names = F, quote = F, col.names = F)

fascicle_names_dep_network_10_percent <- fascicle_volumes_dep_network$fascicle[which(fascicle_volumes_dep_network$inDepMask == 2)] #n=15
#write.table(fascicle_names_dep_network_10_percent,"/Users/eballer/BBL/msdepression/templates/dti/HCP_YA1065_tractography/fascicle_names_dep_network_10_percent.csv", row.names = F, quote = F, col.names = F)

fascicle_names_dep_network_5_percent <- fascicle_volumes_dep_network$fascicle[which(fascicle_volumes_dep_network$prop_in_mask >= 0.05)] #n=24

@@

##########################################################
#             Merge with data_empi              #
##########################################################
df_demo_and_fascicles <- merge(data_empi_acc_f_phq, fascicle_proportions, by = c("EMPI", "EXAM_DATE")) #n = 2962

##########################################################
#                 Histos                        #
##########################################################
##### Histograms for PHQ2/9 healthy and depressed

hist(df_demo_and_fascicles$PHQ.2[which(!is.na(df_demo_and_fascicles$PHQ.2))], main = paste0("PHQ-2 In Good Mimosa Group : n = ", length(which(!is.na(df_demo_and_fascicles$PHQ.2)))), xlab = "PHQ-2", breaks = 8)

hist(df_demo_and_fascicles$PHQ.9[which(!is.na(df_demo_and_fascicles$PHQ.9))], main = paste0("PHQ-9 In Good Mimosa Group : n = ", length(which(!is.na(df_demo_and_fascicles$PHQ.9)))), xlab = "PHQ-9", breaks = 8)

phq2_and_dep <- df_demo_and_fascicles %>% filter(depGroupVar == 2) %>% drop_na(PHQ.2) 
phq9_and_dep <- df_demo_and_fascicles %>% filter(depGroupVar == 2) %>% drop_na(PHQ.9) 
phq2_and_healthy <- df_demo_and_fascicles %>% filter(Has.depdx == 0) %>% drop_na(PHQ.2) 
phq9_and_healthy <- df_demo_and_fascicles %>% filter(Has.depdx == 0) %>% drop_na(PHQ.9)

hist(phq2_and_dep$PHQ.2, main = paste0("PHQ-2/Depressed In Good Mimosa Group : n = ", dim(phq2_and_dep)[1]), xlab = "PHQ-2", breaks = 8)

hist(phq9_and_dep$PHQ.9, main = paste0("PHQ-9/Depressed In Good Mimosa Group : n = ", dim(phq9_and_dep)[1]), xlab = "PHQ-9", breaks = 8)

hist(phq2_and_healthy$PHQ.2, main = paste0("PHQ-2/Healthy In Good Mimosa Group : n = ", dim(phq2_and_healthy)[1]), xlab = "PHQ-2", breaks = 8)
hist(phq9_and_healthy$PHQ.9, main = paste0("PHQ-9/Healthy In Good Mimosa Group : n = ", dim(phq9_and_healthy)[1]), xlab = "PHQ-9", breaks = 8)

######## Histograms for depression network
#this histogram helps contextualize the number of streamlines with varying degrees of proportions affected
hist(x=fascicle_volumes_dep$prop_in_mask, main = "Histogram Affected Streamlines in Dep Net", xlab = "Proportion Streamlines per Fascicle Affected", breaks = 20)

@@

make_demographics_table_ms(data_frame = data_empi_acc_f_phq)

>>>>

```{r read_in_df}
homedir <- "/Users/eballer/BBL/msanxiety/"

#open Rds
data_2023_pull <- readRDS(paste0(homedir, "/data/data_2023_pull_with_curated_icd10_codes.rds"))
```

```{r read_in_mimosa_qc}
#This section checks how many unique individuals had QA data, and then how many met MIMoSA QC. Some additional people lost at registration.

#1 Total unique individuals who also have MS - 890
#2 Total unique who had good MIMoSA - 786
#3 Total unique with bad MIMoSA - 890-786 = 104
#4 Total lost in streamline filtering - 3
#5 Final N going into inclusion/exclusion - 783

#read in mimosa data
mimosa_qc <- read.csv("/Users/eballer/BBL/msdepression/data/melissa_martin_files/csv/mimosa_dataframe", sep = ",", header = FALSE) #n=2840

#preprocess data frame to extract dates, empis, and score
empis_with_qc <- mimosa_qc %>% 
  mutate(EMPI = as.factor(gsub(x = V5, pattern = ".*sub-(.*)/ses-.*/.*", replacement="\\1", perl = TRUE))) %>%
  mutate(EXAM_DATE = gsub(x=V5, pattern = ".*/ses-(.{8})/.*", replacement = "\\1\\2\\3")) %>% 
  mutate(score = V3)

#merge with whole data frame to make sure we exclude people w/out MS
mimosa_qc_df <- merge(data_2023_pull, empis_with_qc, by = c("EMPI", "EXAM_DATE")) #n = 3592

#get list of unique patients, n = 890
mimosa_qc_unique <- mimosa_qc_df %>%
  group_by(EMPI) %>%
  arrange(EXAM_DATE) %>%
  slice(1) %>%
  ungroup #n=890

#get list of unique patients with good QC, n = 867
mimosa_unique_score_gre_75 <- mimosa_qc_unique %>%
  filter(score >= 75) %>%
  group_by(EMPI) %>%
  arrange(EXAM_DATE) %>%
  mutate(EXAM_DATA = as.character(EXAM_DATE)) %>%
  mutate(EMPI = as.character(EMPI)) %>%
  dplyr::select(EMPI, EXAM_DATE, score) %>%
  slice(1) %>%
  ungroup #n=786
```

```{r merge_df_with_mimosa}

#n=867
df_people_with_good_mimosa <- data_2023_pull %>% 
  group_by(EMPI, EXAM_DATE) %>%
  slice(1) %>%
  left_join(mimosa_unique_score_gre_75, by = c("EMPI", "EXAM_DATE")) %>%
  filter(!is.na(score))
```

```{r read_in_fascicle_info}
##########################################################
#             Read in Fascicle info             #
##########################################################

fascicle_proportions <- read.csv(paste0("/Users/eballer/BBL/msdepression//results/fascicle_volumes_all_subjects_roi_n2336.csv"), header = T, sep = ",") #n = 2336
fascicle_proportions <- fascicle_proportions %>%
  mutate(EMPI = as.character(EMPI)) %>%
  mutate(EXAM_DATE = as.character(EXAM_DATE))

@@

fascicle_names_no_cranial_nerves <- fascicle_names[grep("CN", fascicle_names, invert = TRUE)] #n=77, drop cranial nerves

#write.table(fascicle_names,"/Users/eballer/BBL/msdepression/templates/dti/HCP_YA1065_tractography/fascicle_names.csv", row.names = F, quote = F, col.names = F)

########
#fascicle volumes in depression network
########
fascicle_volumes_dep <- read.csv("/Users/eballer/BBL/msdepression/results/streamline_volume_within_dep_network.csv", header = T, sep = ",")

@@

#write.table(fascicle_volumes_dep_network, paste0(homedir,  "/templates/dti/HCP_YA1065_tractography/fascicle_names_dep_network_grouping.csv"), sep = ",", col.names=T)#, row.names = F, quote = F, col.names = T)

@@

@@

##########################################################
#             Merge with data_empi              #
##########################################################

 #n = 823
df_demo_and_fascicles <- df_people_with_good_mimosa %>% 
  left_join(fascicle_proportions, by = c("EMPI", "EXAM_DATE")) %>%
  filter(!is.na(Has.MS.Provider))

#Compare disease burden by anxiety comorbidity
```{r compare_disease_burden_by_anxiety_comorbidity}

#lm
fascicle_number_of_anxiety_dx_lm <- lapply(fascicle_names_no_cranial_nerves, function(x) 
{
  lm(substitute(i ~ number_of_anxiety_comorbidities_extended, list(i = as.name(x))), data = df_demo_and_fascicles)
})
names(fascicle_number_of_anxiety_dx_lm) <- fascicle_names_no_cranial_nerves
fascicle_number_of_anxiety_dx_anova <- lapply(fascicle_number_of_anxiety_dx_lm, anova)

#fdr corrected
fascicle_number_of_anxiety_dx_anova_fdr <- fdr_anova_generic(fascicle_number_of_anxiety_dx_anova, 1)
print(fascicle_number_of_anxiety_dx_anova_fdr)

```
#Has depression AND anxiety diagnoses
```{r compare_disease_burden_by_dep_AND_anxiety_comorbidity}

#lm
fascicle_depANDanx_dx_lm <- lapply(fascicle_names_no_cranial_nerves, function(x) 
{
  lm(substitute(i ~ Has.depANDanx.dx, list(i = as.name(x))), data = df_demo_and_fascicles)
})
names(fascicle_depANDanx_dx_lm) <- fascicle_names_no_cranial_nerves
fascicle_depANDanx_dx_anova <- lapply(fascicle_depANDanx_dx_lm, anova)

#fdr corrected
fascicle_depANDanx_dx_anova_fdr <- fdr_anova_generic(fascicle_depANDanx_dx_anova, 1)
print(fascicle_depANDanx_dx_anova_fdr)

#compare disease burden by psychiatric comorbidity
```{r compare_disease_burden_between_anxiety_and_no_psych_dx}

#lm
fascicle_number_of_psychiatric_dx_lm <- lapply(fascicle_names_no_cranial_nerves, function(x) 
{
  lm(substitute(i ~ number_of_psychiatric_comorbidities_extended, list(i = as.name(x))), data = df_demo_and_fascicles)
})
names(fascicle_number_of_psychiatric_dx_lm) <- fascicle_names_no_cranial_nerves

fascicle_number_of_psychiatric_dx_anova <- lapply(fascicle_number_of_psychiatric_dx_lm, anova)

#fdr corrected
fascicle_number_of_psychiatric_dx_anova_fdr <- fdr_anova_generic(fascicle_number_of_psychiatric_dx_anova, 1)
print(fascicle_number_of_psychiatric_dx_anova_fdr)
```

# Yeo network analysis attempts
###Read in fascicle info
```{r read_in_fascicle_info}
##########################################################
#             Read in Fascicle info             #
##########################################################
########
#fascicle volumes in yeo networks
########

#read in file with proportions
#fascicle_volumes_yeo_7 <- read.csv(paste0(homedir, "results/yeo_7_network_overlap_proportions.txt"), header = T, sep = ",", na.strings="0")

#make titles nice
#names(fascicle_volumes_yeo_7) <-gsub("prop_in_mask_", "prop_", names(fascicle_volumes_yeo_7))

#make Nas into 0
#fascicle_volumes_yeo_7[is.na(fascicle_volumes_yeo_7)]=0

#read in file with binarized yes/no 
fascicle_volumes_yeo_7_binarized <- read.csv(paste0(homedir_msdep, "results/yeo_7_network_overlap_proportions_binarized.txt"), header = T, sep = ",")

#drop cranial nerves
fascicle_volumes_yeo_7_binarized_no_cranial_nerves <- fascicle_volumes_yeo_7_binarized %>% filter(!grepl('CN', fascicle))

#check that order of df is same as yeo networks
new_df <- gsub(pattern = "_vol", replacement = "", x = names(df_demo_and_fascicles_prop_and_vol[grep(pattern = "_vol", x = names(df_demo_and_fascicles_prop_and_vol))]))

combined_df <- data.frame(cbind(new_df, fascicle_volumes_yeo_7_binarized_no_cranial_nerves$fascicle))
names(combined_df) <- c("df", "yeo")

#they match
combined_df$match <- combined_df$df == combined_df$yeo
```

### Calc healthy volume of yeo networks
```{r calc_healthy_volume_of_yeo_networks}
#multiply with the binary yeo maps to get absolute volumes of healthy
fascicle_volumes_yeo_7_full_volumes <- fascicle_volumes_yeo_7_binarized_no_cranial_nerves %>% 
  tibble::remove_rownames() %>%
  tibble::column_to_rownames(var = 'fascicle') %>%
  mutate(across(everything()) * fiber_volume_values_no_cranial_nerves$volume_full) %>%
  rename_with(~ str_c(.x, "_vol"), everything())

#sum the volumes of each network to get total. 
sums_of_volume_of_healthy_fascicles_by_network <- fascicle_volumes_yeo_7_full_volumes %>%
  summarise(across(everything(), ~ sum(., na.rm = TRUE)))

###Calculate injury to networks for each subject
```{r calculate_injury_to_network}

#grab names of networks, get rid of first element which is "fasiscle"
yeo_network_names <- names(fascicle_volumes_yeo_7_binarized_no_cranial_nerves)[-1]

df_w_yeo_volume_info <- dep_and_healthy_groups_for_ICD_analysis_with_hydra_prop %>%
  filter(Hydra_k2 != -1)  %>% 
 # mutate(total_fascicle_vol_lost = rowMeans(across(all_of(fascicle_names_w_vol_suffix)))) %>%
  group_by(EMPI) %>% 
  arrange(EXAM_DATE) %>% 
  slice(1) %>% 
  dplyr::select(EMPI, Hydra_k2, ends_with("_vol")) %>%
  ungroup()

#add columns to data frame for each network
df_w_yeo_volume_info[,yeo_network_names] = 0

#this has been checked with eyeballs and works. You CAN trust the df_w_yeo_vol_info
for (network in yeo_network_names) {

  for (subj in 1:dim(df_w_yeo_volume_info)[1]) {
      # multiply the binary vector of the network with the fascicle volumes of injury, sum, and store in data frame
      text_to_eval <- paste0("df_w_yeo_volume_info$", network,"[", subj,"]=sum(fascicle_volumes_yeo_7_binarized_no_cranial_nerves$", network,
      "*df_w_yeo_volume_info[",subj,",2:(dim(df_w_yeo_volume_info)[2] - 7)])")
      eval(parse(text=text_to_eval))
  }
}

### going to add 7 more columns to the data frame that contain the volume of injury/total healthy volume
df_w_yeo_volume_info_with_prop_of_network <- tibble(df_w_yeo_volume_info) %>% 
  mutate(across(all_of(yeo_network_names), 
    .names = "{.col}_prop_net"

  ))

#go through each subject, and divide their total volume of injury within each network by the overall volume of the network
for (subj in 1:dim(df_w_yeo_volume_info_with_prop_of_network)[1]){
  df_w_yeo_volume_info_with_prop_of_network[subj, grepl(x = names(df_w_yeo_volume_info_with_prop_of_network), pattern = "_prop_net")] = df_w_yeo_volume_info_with_prop_of_network[subj, grepl(x = names(df_w_yeo_volume_info_with_prop_of_network), pattern = "_prop_net")]/sums_of_volume_of_healthy_fascicles_by_network
}

@@

make_demographics_table_ms(data_frame = data_2023_pull)
